{\bf \Large The Power of Context in Decision-Making and Recommendations}

### 2023-08-03 draft.tex
final version submission

    This dissertation comprises of four studies investigating the context effects and applies the knowledge to enhance recommender systems in online marketplaces. I differentiate between two types of context. The first type is intricately tied to the alternatives present within a choice set. The second type of context originates from factors external to the choice set. The findings of the dissertation indicate that internal context effects in choice settings which were primarily studied and observed in experimental settings before are also present and detectable in field settings. Moreover, the study investigating the external context finds the positive relationship between tools that enable users to signal their preferences to recommender systems and their adoption. The empirical applications of the chapters comprising this thesis rely on four distinct and unique datasets, the largest being an observational data from a field setting and the remaining three coming from an experimental setting.

    The first chapter applies a computational decision making model to a substantial real-world choice dataset, making the first application of this magnitude. Findings indicate that context influences individuals' choices. The study suggest that online marketplaces could utilize such models to gain further insights into how the composition of the choice set and interplay among different options affect consumer decisions. The key feature of this computational model is that it identifies context in aggregate. Understanding the underlying components of context can foster a more profound comprehension, offering more granular insights into how each of them individually influence the choice behaviors of individuals when co-existing with other effects.

    Building on the results of the first chapter, the second chapter extends beyond the traditional studies of context, by developing a methodology that aims to disentangle its three main components known as attraction, compromise and similarity from one another. This methodology is designed to be versatile, able to cope with diverse choice settings with varying sizes. Data sourced from both online and offline environments can be successfully integrated into this framework, enhancing its scope and applicability. Insights derived from this study offer potential to fine-tune online product offers and recommender systems. One prominent application of this approach lies in navigating the 'cold-start' problem faced digital marketplaces. It advocates the design of consideration sets that entail empirically similar alternatives, fostering effective decision-making in digitally crowded spaces. This innovative approach represents a stepping stone for future research to extract and utilize contextual information from choice sets. Hence, it contributes to the development of recommender system designs and to the deeper understanding of heterogeneous nature of consumer choice dynamics.

    The third study presents a novel approach to addressing the user-side cold-start problem in recommender system design. It applies the findings of two-stage choice process observed in individuals to contextual information extracted from the choice set via clustering to generate consideration sets. The approach proves to be robust to clustering method and different heuristics applied. Its findings pave the way for investigating context effects arising from outside the choice sets, namely the preferences of individuals and the tools that allow them to signal their preferences to recommender systems. This is argued to be one of the main mechanisms to create more effective recommender systems.

    The final study investigated the effect of user control on recommender system acceptance using the Technology Acceptance Model as a theoretical framework. This study found that easy-to-use recommender systems were perceived as more useful by users and resulted in a higher intention to use them. However, different control mechanisms had varying impacts on user experience. The findings propose an exciting opportunity for future research in designing more effective recommender systems, including the potential integration of large language models (LLMs) for better user interaction and recommendation transparency.

    This thesis demonstrates the existence of context effects in multi-attribute, multidimensional settings and develops methodologies of enhancing recommender systems design with context effects. Additionally, this study investigates external context, namely, the tools that allows users to express their preferences and how to make recommender systems better through them.

@@

Embarking on this scholarly journey has been both humbling and transformative, marked by countless days of meticulous study, intense deliberation, and profound discoveries. Indeed, the journey was punctuated with hurdles and moments of self-doubt. However, every trough was succeeded by a crest, each stumble by an ascent, and every challenge served as an opportunity. The path was steep, but it was these uphill struggles that honed my perspective, expanded my horizons, and ignited my passion for knowledge. The lessons learned, the triumphs celebrated, and the sense of accomplishment far outweighed the trials faced. Reflecting on these uplifting moments, I joyfully acknowledge those who accompanied me on this journey and made it not only possible but immensely rewarding.

Firstly, I extend my deepest gratitude to Zakaria Babutsidze, my PhD supervisor. He has been a great mentor and a source of inspiration for me. His honesty and straightforwardness challenged me to think critically and creatively, shaping me into someone better day by day. I am indebted to Francesco Castellanetta, our PhD program director, for his guiding hand and unwavering faith in my capabilities. I appreciate our long and intriguing discussions with Ludovic Diabaggio that transcended academia. I am grateful to William Rand for his invaluable insights and co-authorship. Daniela Iubatti, your informal talks and unwavering support have been a source of inspiration. I appreciate Diego Zunino for his earnest and invaluable guidance. Bruno Cirillo deserves special mention; our albeit slightly formal discussions have significantly shaped my work. My heartfelt appreciation goes to Renata Kaminska. Her support and intuitive understanding of my thoughts were indispensable in my pursuits. I am grateful to Benjamin Montmartin; our separate exchanges about wine (we were in France in the end) and statistics enriched my perspectives immeasurably. To Valerio Incerti, our creative discourses, even about less creative ideas were a constant source of inspiration. Lapo Mola’s precious feedback motivated me to pursue excellence, and I value that wholeheartedly.

Aytan, my beloved wife and best friend, your unwavering support and guidance have been my Polaris guiding me through the darkest nights and taught me to only settle for infinity and beyond. Ecem Delicik, my dear friend, I am grateful for the shared stories, moments, laughter, and silence that have strengthened our friendship. The moment you realized I was speaking to you in Turkish is unforgettable (I am sure, you still remember it too). Johanna Deperi, our shared moments and long talks were invaluable for self-reflection, joy, and sometimes, ``justified sadness'' throughout this journey. Artyom Yepremyan, our discussions helped us conquer the PhD mountain. I will forever cherish the times when I needed a hand, and you extended yours. Manon Desjardins, your constant support despite the towering stack of papers on your desk will always be appreciated. Mehdi Ibrahim, I am grateful for our coffee talks and the irresistible desserts from Morocco. Teymur Mardaliyev, your friendship and coding prowess have been invaluable; only you could convince me to settle for JavaScript there.

Finally, no amount of words can fully express my gratitude to my parents. Dad, your inquisitive nature has been my guiding star, instilling in me the curiosity that set me on this scholarly path. Your enthusiasm for the uncharted has been infectious, leading me to question, explore, and discover. Mom, you have been my anchor, reminding me of my roots and my identity amidst the academic whirlwind. Your unwavering encouragement and steadfast belief have reinforced my sense of self, allowing me to face challenges with courage and resilience. You taught me to stay grounded and true to myself, no matter how far I travel in the pursuit of knowledge. Your lessons continue to shape my journey, helping me navigate both personal and academic hurdles with grace and fortitude.

@@

The main objective of this thesis is to make contribution towards understanding the choice context and applying it to improve recommender systems. Thesis makes effort by identifying context effects arising from the choice set and developing a methodology to implement this information in recommender systems design. Furthermore, it analyzes the effectiveness of user control mechanisms, which allow individuals to inform the system about their preferences and also amend them.

Recommender systems have become a crucial ally in the vast landscape of digital world. They help individuals to navigate through the plethora of alternatives and find what they are looking for. They achieve this by using sophisticated algorithms that analyze the wide space of items, users and their interactions with each other. System designers and businesses thrive to maximize the accuracy of recommendations, meaning they want to maximize the consumption of the recommendations. However, recent research stream shows that, accuracy is not the only way to make recommender systems effective. Other factors like diversity, serendipity are important too\citep{kaminskas2016diversity}. Moreover, the black-box nature of the algorithms do not allow us to understand the reasons behind a particular user's choice behaviour \citep{kotkovSurveySerendipityRecommender2016, samih2021exmrec2vec}. Hence, more and more interest is driven towards the understanding and utilizing the context around the particular choice \citep{adomavicius2005toward}.

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates on and uses two of them. The first one posits that choice context \footnote{For the remainder of the dissertation, I will follow the previous literature \citep{truebloodEtAl13} and refer to this definition of choice context as context effects.} is ``the availability and the nature of choice alternatives'' \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Then, context is also referred to as ``the time and content of the choice, the location or socio-demographic characteristics of the decision-maker...'' \citep{adomavicius2011context}\footnote{To distinguish between these two definitions I will refer to this definition as external context throughout this dissertation.}

However, modeling for context effects mathematically was a challenge because the existing models suffered from independence of irrelevant alternatives criterion \citep{luce59}, meaning these models treated each alternative in isolation. This has shifted the scholar's interest towards computational models \citep{usher2001time, roe2001multialternative, trueblood2014multiattribute, noguchi2018multialternative}. Yet, none of these models has been applied to real-world choice data and as a result, their applicability to field data remained as an open question. I address this gap in the first study where I apply a computational model to a field data of high heterogeneity among dimensions. Building on my results in my second study I delve into the choice modeling more and develop a methodology to account for three main context effects studied in the literature \citep{truebloodEtAl13}.

When new users or items are introduced to recommender systems, they fail to function as intended because of the lack of information their algorithms need. In such cases, even when they do provide recommendations, recommendations are far from personalized \citep{lika2014facing}. It is considered to be a key challenge in recommender system design \citep{park2009pairwise}. Previous literature has applied various methods to overcome this problem, including asking users to rate some items, share their preferences among others \citep{guy2009personalized, aharon2013off, bykau2013coping, saveski2014item}. However, the limitations of these approaches are that they ignore the context effects and concentrate on scenarios where information scarcity is temporary. Using the findings from my second study and combining it with decision-making and choice modeling literature I propose an innovative solution to address continuous information scarcity issue in recommender system design by utilizing contextual information of the menu to generate relevant choice sets through a two-step choice modeling method.

Research agrees on importance of metrics outside the boundaries of accuracy for recommender systems \citep{kaminskas2016diversity}. It is in the best interest of online marketplaces to provide users with recommendations which are not only in the category ``exactly what I want'', but also ``I never thought I would have liked this one'' \citep{kotkovSurveySerendipityRecommender2016}. This can not only boost sales \citep{songWhenHowDiversify2019}, but also increase the satisfaction of users \citep{knijnenburgExplainingUserExperience2012}. Albeit, it was observed that the preferences of individuals are not stable and tend to change and system designers have proposed various tools to enable users signal their preference shifts \citep{bostandjiev2012tasteweights,hijikata2012relation}. However, to better understand it is necessary to study those tools in isolation from the context arising from the choice set and concentrate purely on external context that arises from the user side, e.g their preferences \citep{adomavicius2011context}. In my last study I address this gap by conducting an online experiment and applying Technology Acceptance Model \citep{davis1985technology} to measure users' acceptance of such tools.

All in all, all four studies aim to contribute to our understanding of context and provide applications of recommender system design using this information.

@@

    Previous computational decision making models that were developed to account for context effects have only been
    studied with an experimental data where only one effect was produced at a time. Using data coming from strictly controlled experimental environments
    hinders the understanding of context effects that occur in real-world choice scenarios where items have multiple dimensions and choice sets have dozens of alternatives. In this chapter
    I apply a computational model that accounts for context effects to an observational data which was not done before. The data comes from an air travel industry and is ideal to study context effects in multi-attribute, multi-alternative choice environments. I first find optimal parameters for computational model using differential evolution algorithm. Then, I complement a traditional choice model with its outputs and assess the significance of its contribution. This chapter contributes to context effect and decision making literature by providing further insights on behavior of computational decision making models in real-world choice data.

@@

Context effects have been extensively studied and demonstrated in various domains, from psychology to marketing \citep{herne1997decoy, soltani2012range, truebloodEtAl13, frederickEtAl14, evangelidisEtAl18, wuConsguner20}. Some recent studies also concluded that multiple context effects may occur at the same time \citep{berkowitsch2014rigorously, noguchi2014attraction}. However, recent studies also have discovered boundary conditions for these effects \citep{liew2016appropriacy, spektor2018good, spektor2019similarity}. Familiarity with the choice domain was found to reduce context effects experienced by individuals \citep{kim2005attraction, sheng2005understanding}.  It was also found that, some conditions may force these effects to completely reverse \citep{cataldo2019comparison}. The findings described above make it necessary to call for a model which could explain these effects.

Logit and Probit models have been traditionally used in choice settings \citep{gensch1979multinomial, kim2017probit}. However, those models can not account for context effects, because they only account for the attributes of the focal option, not taking into account the attributes of the other options in the choice set. Tversky has proposed a model of elimination by aspects which could account for similarity effect \citeyearonly{tversky1972elimination}. The foundation of the model is attention switching of individuals between alternatives and attributes and their comparisons. Once receiving attention, attribute value of a given alternative is compared to a pre-determined threshold value by the individual and if failing to meet the threshold, that alternative is eliminated from the decision. The step is then  continued with another attribute until final decision is made. Another model which was proposed by Tversky and Simonson could account for compromise effect \citeyearonly{tverskySimonson93}. That model posited that, alternatives are compared based on a weighted sum of attribute values and a local context comprising of binary comparisons among alternatives. However, these two models could not successfully account for all three effects. Despite their drawbacks, the sequential decision making and attention based mechanisms in these models laid the foundations of many upcoming computational choice models \citep{bhatia2013associations}.

In the last three decades, researchers have developed many computational models which account for context effects: Multialternative Decision Field Theory (MDFT) \citep{roe2001multialternative}, Leaky competing
accumulator \citep{usher2001time}, Multiattribute linear ballistic accumulator \citep{trueblood2014multiattribute}, Multialternative Decision by Sampling  \citep{noguchi2018multialternative} are part of them. Some of these models have been extensively tested and studied, while others are relatively new, hence have not received much attention from scholars \citep{truebloodEtAl13}. However, these studies have been performed with experimental data \citep{trueblood2014multiattribute,  berkowitsch2014rigorously, evans2019response, busemeyer2019cognitive}. Research has proven that the behavior of individuals in laboratory choice environment is different from real-world choice environment \citep{hogarth1989risk}. Hence, the applicability of such models to field data is an uninvestigated avenue, because no previous study has been done where a computational model was been applied to a real-world observational data. The plan is to address this gap by applying Multialternative Decision by Sampling (MDbS) model proposed by Noguchi and Stewart \citeyearonly{noguchi2018multialternative} to a field data from airfare booking domain. This would allow me to assess the applicability level of this model complex field-data.

Also, applying MDbS to observational data would allow me to statistically asses the significance of the contribution of this model's ability to account for context effect. Instead of testing this model against established choice models, I will attempt to complement them with MDbS. To do this, random effect Probit model as a variation of Probit family will be used and will be augmented with MDbS output. The Probit model is chosen as it does not explicitly assume IIA unlike family of Logit models. To validate my results further I will apply the same methodology and analysis to an experimental data.

The reasons behind choosing MDbS are twofold. Firstly, it is relatively new when compared to other models, hence it has not been further investigated before. Secondly, it is more robust and can account for wider range of context effects (than other models) known to the literature \citep{noguchi2018multialternative}.

@@

We make choices all the time. Imagine the time when you went to see a movie you have been waiting for some time ago and decided to grab some popcorn before entering. You might have seen something like in the figure \ref{fig:decoyPopcornExample}, although prices may be higher these days. You are puzzled at first, but reminded that the movie about to start, so you better hurry up and choose one.

The small one feels not enough for a 90 minute marathon. Then, there is a middle one which seems like an okay option, at first. When you notice that big box you immediately forget about the small one you saw moments ago. You start looking at sizes and prices of middle and large box and think: Well, that's easy. Big box seems like the way to go here, considering their prices are almost equal.

@@

If this situation is familiar to you, then you have experienced so called context effect which can bu understood as ``the composition  and the nature of the choice set, availability of various options in it'' \citep{tversky1972elimination, huberPuto83}. For the long time, our understanding of the choice did not extend over the borders of two related principles. First one is, independence of irrelevant alternatives (IIA) which states that when having a choice between two two options $A$ and $B$ if a person prefers $A$ for example, regardless of adding a third option $C$ to this choice set, that person's preference must be unaltered \citep{luce59}. Second one is regularity principle which states that, the choice probability of option $A$ can not increase by introduction of option $C$ \citep{luce59}.

However, research has concluded that the context of the choice set and options in it have substantial effect on how people make choices. This effect has been studied extensively for the past five decades by many economists, marketing scholars and psychologists \footnote{See Dowling et al. \citeyearonly{dowlingEtAl20} and Lichtenstein \citeyearonly{lichtenstein2006construction} for more comprehensive review.} \citep{ kahnemanTversky79, simonson89, tverskySimonson93, lichtenstein2006construction, dowlingEtAl20}. Most of the research has focused on three context effects, also referred to as ``big three'': attraction, compromise and similarity \citep{howes2016contextual}.

To better understand these three context effects, lets think of a hypothetical choice set where options differ along two dimensions: Dimension 1 and Dimension 2. We first start with a choice set consisting of two options: $A$ which has the values 20 and 80; $B$ which has values 80 and 20 that figure \ref{fig:binaryChoiseSet} depicts.

@@

People who prefer Dimension 1 would choose $B$ whereas $A$ will be chosen by individuals preferring Dimension 2. This can be described as equation \ref{eq:onlyTwoOptions} below.

@@

Where $P(A|A,B)$ corresponds to the probability of choosing $A$ given choice set $A,B$. Same goes for $P(B|A,B)$.

@@

Now, lets add a third option to this choice set, option $D_A$ to create one variation and $D_B$ to create the second variation of the ternary choice set. Both added options have lower values in both dimensions when compared to $A$ and $B$ respectively. Huber created this type of scenario and has found what he has then called ``attraction effect'' \citeyearonly{huberEtAl82}. Attraction effect, which is also known in the literature as asymmetric dominance effect is a consistent violation of the regularity principle mentioned earlier. He suggested that when having a choice set consisting of options $A$ and $B$ the relative probability of choosing option $A$ can be increased if one adds a third option with characteristics of $D_A$ to the same choice set \citeyearonly{huberEtAl82}.

@@

Figure \ref{fig:attractionEffect} shows those options and their respective values in each dimension. One can observe that option $A$ and $B$ are located in two different ends of the choice space. Option $D_A$ is inferior to option $A$ in both dimensions and $D_B$ is inferior to $B$. With attraction effect in place equation  \ref{eq:onlyTwoOptions} will change into \ref{eq:attractionProbability}.

@@

\State Initialize population of $P$ from parameter space $S$

>>>>

    This dissertation comprises four studies investigating context effects and applying the knowledge to enhance recommender systems in online marketplaces. I differentiate between two types of context. The first type is intricately tied to the alternatives present within a choice set. The second type of context originates from factors external to the choice set. The findings of the dissertation indicate that internal context effects in choice settings, which were primarily studied and observed in experimental settings before, are also present and detectable in field settings. Furthermore, the study investigating the external context finds a positive relationship between tools that enable users to signal their preferences to recommender systems and their adoption. The empirical applications of the chapters comprising this thesis rely on four distinct and unique datasets, the largest being observational data from a field setting and the remaining three coming from an experimental setting.

    The first chapter applies a computational decision-making model to a substantial real-world choice dataset, making the first application of this magnitude. Findings indicate that the context influences individuals' choices. The study suggests that online marketplaces could utilize such models to gain further insights into how the composition of the choice set and the interaction among different options affect consumer decisions. The key feature of this computational model is that it identifies context in aggregate. Understanding the underlying components of context can foster a more profound comprehension, offering more granular insights into how each of them individually influences the choice behaviors of individuals when coexisting with other effects.

    Building on the results of the first chapter, the second chapter extends beyond the traditional studies of context by developing a methodology that aims to disentangle its three main components known as attraction, compromise, and similarity from one another. This methodology is designed to be versatile, able to cope with diverse choice settings of varying sizes. Data sourced from both online and offline environments can be successfully integrated into this framework, enhancing its scope and applicability. The insights derived from this study offer the potential to fine-tune online product offers and recommender systems. One prominent application of this approach lies in navigating the 'cold-start' problem faced digital marketplaces. It advocates the design of consideration sets that entail empirically similar alternatives, fostering effective decision-making in digitally crowded spaces. This innovative approach represents a stepping stone for future research in extracting and utilizing contextual information from a set of choices. Therefore, it contributes to the development of recommender system designs and to a deeper understanding of the heterogeneous nature of the dynamics of consumer choice.

    The third study presents a novel approach to addressing the user-side cold start problem in recommender system design. It applies the findings of the two-stage choice process observed in individuals to contextual information extracted from the choice set by clustering to generate consideration sets. The approach proves to be robust to the clustering method and different heuristics applied. Its findings pave the way for investigating context effects arising from outside the choice sets, namely, the preferences of individuals and the tools that allow them to signal their preferences to recommender systems. This is argued to be one of the main mechanisms to create more effective recommender systems.

    The final study investigated the effect of user control on recommender system acceptance using the Technology Acceptance Model as a theoretical framework. This study found that easy-to-use recommender systems were perceived by users as more useful and resulted in greater intention to use them. However, different control mechanisms had varying impacts on user experience. The findings propose an exciting opportunity for future research to develop more effective recommender systems, including the potential integration of large language models (LLMs) for better user interaction and recommendation transparency.

    This thesis demonstrates the existence of context effects in multiattribute, multidimensional settings and develops methodologies of enhancing recommender systems design with context effects. Additionally, this study investigates the external context, namely, the tools that allow users to express their preferences and how to make recommender systems better through them.

@@

Embarking on this scholarly journey has been both humbling and transformative, marked by countless days of meticulous study, intense deliberation, and profound discoveries. Indeed, the journey was punctuated with hurdles and moments of self-doubt. However, every trough was succeeded by a crest, each stumble by an ascent, and every challenge served as an opportunity. The path was steep, but these uphill struggles honed my perspective, expanded my horizons, and ignited my passion for knowledge. The lessons learned, the triumphs celebrated and the sense of accomplishment far outweighed the trials faced. Reflecting on these uplifting moments, I joyfully acknowledge those who accompanied me on this journey and made it not only possible, but also immensely rewarding.

Firstly, I extend my deepest gratitude to Zakaria Babutsidze, my Ph.D. supervisor. He has been a great mentor and a source of inspiration for me. His honesty and straightforwardness challenged me to think critically and creatively, shaping me into someone better day by day. I am indebted to Francesco Castellanetta, our Ph.D. program director, for his guiding hand and unwavering faith in my capabilities. I appreciate our long and intriguing discussions with Ludovic Diabaggio that transcended academia. I am grateful to William Rand for his invaluable insights and coauthorship. Daniela Iubatti, your informal talks and unwavering support have been a source of inspiration. I appreciate Diego Zunino for his honest and invaluable guidance. Bruno Cirillo deserves special mention; our albeit slightly formal discussions have significantly shaped my work. My heartfelt appreciation goes to Renata Kaminska. Her support and intuitive understanding of my thoughts were essential in my pursuits. I am grateful to Benjamin Montmartin; our separate exchanges about wine (we were in France at the end) and statistics enriched my perspectives immeasurably. To Valerio Incerti, our creative discourses, even about less creative ideas, were a constant source of inspiration. Lapo Mola's valuable feedback motivated me to pursue excellence, and I value that wholeheartedly.

Aytan, my beloved wife and best friend, your unwavering support and guidance have been my Polaris, guiding me through the darkest nights and teaching me to only settle for infinity and beyond. Ecem Delicik, my dear friend, I am grateful for the shared stories, moments, laughter, and silence that have strengthened our friendship. The moment you realized that I was speaking to you in Turkish is unforgettable (I am sure you still remember it, too). Johanna Deperi, our shared moments and long conversations were invaluable for self-reflection, joy, and sometimes ``justified sadness'' throughout this journey. Artyom Yepremyan, our discussions helped us conquer the PhD mountain. I will forever cherish the times when I needed a hand, and you extended yours. Manon Desjardins, your constant support despite the towering stack of papers on your desk will always be appreciated. Mehdi Ibrahim, I am grateful for our coffee talks and the irresistible desserts from Morocco. Teymur Mardaliyev, your friendship and coding prowess have been invaluable; only you could convince me to settle for JavaScript there.

Finally, no words can fully express my gratitude to my parents. Father, your inquisitive nature has been my guide, instilling in me the curiosity that led me on this scholarly path. Your enthusiasm for the unexplored has been infectious, leading me to question, explore, and discover. Mother, you have been my anchor, reminding me of my roots and my identity in the academic whirlwind. Your unwavering encouragement and steadfast belief have reinforced my sense of self, allowing me to face challenges with courage and resilience. You taught me to stay grounded and true to myself, no matter how far I travel in the pursuit of knowledge. Your lessons continue to shape my journey, helping me navigate both personal and academic hurdles with grace and fortitude.

@@

The main objective of this thesis is to make a contribution towards understanding the choice context and applying it to improve recommender systems. The thesis makes an effort to identify context effects arising from the choice set and to develop a methodology to implement this information in the design of recommender systems. Furthermore, it analyzes the effectiveness of user control mechanisms, which allow individuals to inform the system about their preferences and also amend them.

Recommender systems have become a crucial ally in the vast landscape of the digital world. They help individuals to navigate through the plethora of alternatives and find what they are looking for. They achieve this by using sophisticated algorithms that analyze the wide space of items, users, and their interactions with each other. System designers and businesses thrive to maximize the accuracy of recommendations, which means they want to maximize the consumption of recommendations. However, recent research streams show that accuracy is not the only way to make recommender systems effective. Other factors such as diversity and serendipity are also important \citep{kaminskas2016diversity}. Moreover, the black-box nature of the algorithms does not allow us to understand the reasons behind a particular user's choice behavior \citep{kotkovSurveySerendipityRecommender2016, samih2021exmrec2vec}. Therefore, more and more interest is directed towards understanding and utilizing the context around the particular choice \citep{adomavicius2005toward}.

The choice context has long received the attention of scholars in many fields, including marketing, psychology, and economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates on and uses two of them. The first one posits that the choice context \footnote{For the remainder of the dissertation, I will follow the previous literature \citep{truebloodEtAl13} and refer to this definition of the choice context as context effects.} is ``the availability and nature of the choice alternatives'' \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Then, context is also referred to as ``the time and content of the choice, the location or sociodemographic characteristics of the decision maker...'' \citep{adomavicius2011context}\footnote{To distinguish between these two definitions I will refer to this definition as external context throughout this dissertation.}

However, modeling for context effects mathematically was a challenge because the existing models suffered from the independence of irrelevant alternative criterion \citep{luce59}, which meant that these models treated each alternative in isolation. This has shifted the scholar's interest towards computational models \citep{usher2001time, roe2001multialternative, trueblood2014multiattribute, noguchi2018multialternative}. Yet, none of these models has been applied to real-world choice data, and as a result, their applicability to field data remained an open question. I address this gap in the first study where I apply a computational model to a field data of high heterogeneity among dimensions. Building on my results in my second study, I have delved deeper into choice modeling and developed a methodology to account for the three main context effects studied in the literature \citep{truebloodEtAl13}.

When new users or items are introduced to recommender systems, they fail to function as intended because of the lack of information their algorithms need. In such cases, even when they do provide recommendations, the recommendations are far from personalized \citep{lika2014facing}. It is considered a key challenge in recommender system design \citep{park2009pairwise}. The previous literature has applied various methods to overcome this problem, including asking users to rate some items, share their preferences among others \citep{guy2009personalized, aharon2013off, bykau2013coping, saveski2014item}. However, the limitations of these approaches are that they ignore the context effects and concentrate on scenarios where information scarcity is temporary. Using the findings from my second study and combining it with decision-making and choice modeling literature, I propose an innovative solution to address continuous information scarcity issue in recommender system design by utilizing contextual information of the menu to generate relevant choice sets through a two-step choice modeling method.

Research agrees on the importance of metrics outside the boundaries of accuracy for recommender systems \citep{kaminskas2016diversity}. It is in the best interest of online marketplaces to provide users with recommendations that are not only in the category ``exactly what I want'', but also ``I never thought I would have liked this'' \citep{kotkovSurveySerendipityRecommender2016}. This can not only boost sales \citep{songWhenHowDiversify2019}, but also increase user satisfaction \citep{knijnenburgExplainingUserExperience2012}. However, it was observed that individual preferences are not stable and tend to change and system designers have proposed various tools to allow users to signal their preference shifts \citep{bostandjiev2012tasteweights,hijikata2012relation}. However, to better understand, it is necessary to study those tools in isolation from the context arising from the choice set and concentrate purely on external context that arises from the user side, e.g., their preferences \citep{adomavicius2011context}. In my last study, I address this gap by conducting an online experiment and applying Technology Acceptance Model \citep{davis1985technology} to measure users' acceptance of such tools.

All in all, the four studies aim to contribute to our understanding of the context and provide applications of the design of recommender systems using this information.

@@

    Previous computational decision making models that were developed to account for context effects have only been studied with an experimental data where only one effect was produced at a time. Using data coming from strictly controlled experimental environments hinders the understanding of context effects that occur in real-world choice scenarios where items have multiple dimensions and choice sets have dozens of alternatives. In this chapter, I apply a computational model that accounts for context effects to an observational data which was not done before. The data comes from an air travel industry and is ideal to study context effects in multiattribute, multialternative choice environments. I first find optimal parameters for the computational model using the differential evolution algorithm. Then, I complement a traditional choice model with its outputs and assess the significance of its contribution. This chapter contributes to context effect and decision-making literature by providing further insights on behavior of computational decision-making models in real-world choice data.

@@

Context effects have been extensively studied and demonstrated in various domains, from psychology to marketing \citep{herne1997decoy, soltani2012range, truebloodEtAl13, frederickEtAl14, evangelidisEtAl18, wuConsguner20}. Some recent studies also concluded that multiple context effects may occur at the same time \citep{berkowitsch2014rigorously, noguchi2014attraction}. However, recent studies have also discovered boundary conditions for these effects \citep{liew2016appropriacy, spektor2018good, spektor2019similarity}. Familiarity with the choice domain was found to reduce the context effects experienced by individuals \citep{kim2005attraction, sheng2005understanding}.  It was also found that some conditions may force these effects to completely reverse \citep{cataldo2019comparison}. The findings described above make it necessary to call for a model which could explain these effects.

Logit and Probit models have been traditionally used in choice settings \citep{gensch1979multinomial, kim2017probit}. However, those models cannot account for context effects because they only account for the attributes of the focal option, not taking into account the attributes of the other options in the choice set. Tversky has proposed a model of elimination by aspects that could explain the similarity effect \citeyearonly{tversky1972elimination}. The foundation of the model is attention switching of individuals between alternatives and attributes and their comparisons. Once attention is received, the attribute value of a given alternative is compared to a predetermined threshold value by the individual, and if the threshold does not meet, that alternative is eliminated from the decision. The step is then continued with another attribute until the final decision is made. Another model proposed by Tversky and Simonson could explain the effect of compromise \citeyearonly{tverskySimonson93}. The theoretical foundations of this model posited that alternatives are compared based on a weighted sum of attribute values and a local context comprising binary comparisons among alternatives. However, these two models were unable to successfully account for all three effects. Despite their drawbacks, the sequential decision making and attention-based mechanisms in these models laid the foundations of many upcoming computational choice models \citep{bhatia2013associations}.

In the last three decades, researchers have developed many computational models which account for context effects. Multialternative Decision Field Theory (MDFT) \citep{roe2001multialternative}, Leaky competing
accumulator \citep{usher2001time}, Multiattribute linear ballistic accumulator \citep{trueblood2014multiattribute}, Multialternative Decision by Sampling  \citep{noguchi2018multialternative} are part of them. Some of these models have been extensively tested and studied, while others are relatively new and therefore have not received much attention from scholars \citep{truebloodEtAl13}. However, these studies have been performed with experimental data \citep{trueblood2014multiattribute,  berkowitsch2014rigorously, evans2019response, busemeyer2019cognitive}. Research has proven that the behavior of individuals in the laboratory choice environment is different from real-world choice environment \citep{hogarth1989risk}. Hence, the applicability of such models to field data is an uninvestigated avenue because no previous study has been done where a computational model was applied to real-world observational data. The plan is to address this gap by applying the Multialternative Decision by Sampling (MDbS) model proposed by Noguchi and Stewart \citeyearonly{noguchi2018multialternative} to a field data from the airfare booking domain. This would allow me to assess the applicability level of this model complex field-data.

Also, applying MDbS to observational data would allow one to statistically assess the significance of the contribution of this model's ability to account for the context effect. Instead of testing this model against established choice models, I will attempt to complement them with MDbS. To do this, the random effect Probit model will be used as a variation of the Probit family and will be augmented with MDbS output. The Probit model is chosen as it does not explicitly assume IIA unlike the family of Logit models. To validate my results further, I will apply the same methodology and analysis to an experimental data.

The reasons for choosing MDbS are twofold. First, it is relatively new when compared to other models, hence it has not been further investigated before. Second, it is more robust and can account for a wider range of context effects (than other models) known to the literature \citep{noguchi2018multialternative}.

@@

We make choices all the time. Imagine the time you went to see a movie you had been waiting for a while and decided to grab some popcorn before entering. You may have seen something like that in the figure \ref{fig:decoyPopcornExample}, although prices may be higher these days. You are puzzled at first, but reminded that the movie is about to start, so you better hurry up and choose one.

The small one feels not enough for a 90 minute marathon. Then, there is a middle one which seems like an okay option at first. When you notice that big box, you immediately forget about the small one you saw moments ago. You start looking at sizes and prices of middle and large boxes and think: Well, that is easy. Big box seems like the way to go here, considering their prices are almost equal.

@@

If this situation is familiar to you, then you have experienced the so-called context effect which can be understood as ``the composition and the nature of the choice set, availability of various options in it'' \citep{tversky1972elimination, huberPuto83}. For a long time, our understanding of the choice did not extend over the borders of two related principles. The first one is independence of irrelevant alternatives (IIA), which states that when having a choice between two two options $A$ and $B$ if a person prefers $A$ for example, regardless of adding a third option $C$ to this choice set, that person's preference must be unaltered \citep{luce59}. The second is the regularity principle, which states that the probability of choice of option $A$ cannot increase by the introduction of option $C$ \citep{luce59}.

However, research has concluded that the context of the choice set and the options in it have a substantial effect on how people make decisions. This effect has been extensively studied over the past five decades by many economists, marketing scholars, and psychologists \footnote{See Dowling et al. \citeyearonly{dowlingEtAl20} and Lichtenstein \citeyearonly{lichtenstein2006construction} for a more comprehensive review.} \citep{ kahnemanTversky79, simonson89, tverskySimonson93, lichtenstein2006construction, dowlingEtAl20}. Most of the research has focused on three context effects, also referred to as ``big three'': attraction, compromise and similarity \citep{howes2016contextual}.

To better understand these three context effects, let us think of a hypothetical choice set where options differ along two dimensions: Dimensions 1 and 2. We first start with a set of options consisting of two options: $A$ that has the values 20 and 80; $B$ that has the values 80 and 20 that figure \ref{fig:binaryChoiseSet} represents.

@@

People who prefer Dimension 1 would choose $B$ whereas $A$ will be chosen by people who prefer Dimension 2. This can be described as equation \ref{eq:onlyTwoOptions} below.

@@

Where $P(A|A,B)$ corresponds to the probability of choosing $A$ given the choice set $A,B$. Same goes for $P(B|A,B)$.

@@

Now, let us add a third option to this choice set, the option $D_A$ to create one variation, and $D_B$ to create the second variation of the ternary choice set. Both added options have lower values in both dimensions compared to $A$ and $B$, respectively. Huber created this type of scenario and has found what he has called the ``attraction effect'' \citeyearonly{huberEtAl82}. The attraction effect, which is also known in the literature as the asymmetric dominance effect, is a consistent violation of the regularity principle mentioned earlier. He suggested that when having a choice set consisting of options $A$ and $B$ the relative probability of choosing option $A$ can be increased if a third option with characteristics of $D_A$ is added to the same set of choices \citeyearonly{huberEtAl82}.

@@

Figure \ref{fig:attractionEffect} shows these options and their respective values in each dimension. It can be seen that option $A$ and $B$ are located at two different ends of the choice space. Option $D_A$ is inferior to option $A$ in both dimensions and $D_B$ is inferior to $B$. With the attraction effect in place, equation \ref{eq:onlyTwoOptions} will change into \ref{eq:attractionProbability}.

@@

\State Initialize population of $P$ from the parameter space $S$
